\mutagen: Reliable Coverage-Guided, Property-Based Testing using Exhaustive Well-Typed Mutations

### 2021-03-31 papers/oopsla-21/main.tex
WIP

There exists a vast literature on fuzzing and property-based testing.
%
In this section we focus on three main topics:
%
automated random data generation using static information,
%
fuzzing using coverage information to guide mutations, and
%
exhaustive bounded testing tools, which inspired the design of \mutagen.

@@

% DRAGEN and DRAGEN2
In Haskell, DRAGEN \cite{dragen} is a meta-programming tool that synthezises
random generators from datatypes definitions, using stochastic models to predict
and optimize their distribution based on a target distribution set by the user.
%
DRAGEN2 \cite{dragen2} extends this idea adding support for generating richer
random data by extracting additional static information like library APIs and
function input patterns from the codebase.
%
% Bolztmann-brain
Similarly, \cite{bendkowski} developed a polinomial tuning mechanism based on
Boltzmann samplers \cite{boltzmann} that synthesizes generators with
approximate-size distributions for combinatorial structures like lists or trees.
%

@@

We consider all these generational approaches to be orthogonal to the ideas
behind \mutagen.
%
In principle, our tool is tailored to improve the performance of poor
automatically derived generators.
%
However, more specialized generators can by used directly by \mutagen, and
combining them with well-typed mutations using perhaps a hybrid technique is an
appealing idea that we aim to address in the future.

@@

% AFL

% AFLfast

% CollAFL

% Criterion

% CSmith, XSmith

% QuickFuzz

Pellentesque dapibus suscipit ligula. Donec posuere augue in quam. Etiam vel
tortor sodales tellus ultricies commodo. Suspendisse potenti. Aenean in sem ac
leo mollis blandit. Donec neque quam, dignissim in, mollis nec, sagittis eu,
wisi. Phasellus lacus. Etiam laoreet quam sed arcu. Phasellus at dui in ligula
mollis ultricies. Integer placerat tristique nisl. Praesent augue. Fusce
commodo. Vestibulum convallis, lorem a tempus semper, dui dui euismod elit,
vitae placerat urna tortor vitae lacus. Nullam libero mauris, consequat quis,
varius et, dictum id, arcu. Mauris mollis tincidunt felis. Aliquam feugiat
tellus ut neque. Nulla facilisis, risus a rhoncus fermentum, tellus tellus
lacinia purus, et dictum nunc justo sit amet elit.

@@

%
% Feat
%
\emph{Feat} \cite{feat} formalizes the notion of \emph{functional enumerations},
For any algebraic type, it synthesizes a bijection between a finite prefix of
the natural numbers and set of increasingly growing values of the input type.
%
Later, this bijection can be traversed exhaustively or, more interestingly,
randomly accesed.
%
This allows the user to easily generate random values in a uniform basis simply
by sampling natural numbers.
%
However, values are enumerated based only on their type definition, so this
technique is not suitable for testing properties with sparse preconditions
expressed elsewhere.

% SmallCheck and LazySmallCheck
\emph{SmallCheck} \cite{smallcheck} is a Haskell tool that follows this idea.
%
\emph{SmallCheck} is an exhaustive testing tool for Haskell that progressively
executes the testing properties against all possible inputs values of within
a certain size bound.
%
Later, \emph{LazySmallCheck} is a lazy variant of \emph{SmallCheck} that uses
lazy evaluation to automatically prune the search space by detecting input
subexpressions that are not evaluated by the testing properties, and hence do
not need to be populated exhaustively with concrete values.

@@

Being exhaustive, these approaches can only be used to enumerate and test a
relatively little space of small inputs.
%
In this work we use exhaustiveness as a way to reliably schedule different
mutants.
%
In contrast to fully-exhaustive testing tools, \mutagen relies initially on
randomness to find interesting mutable candidates.
%
Later, enumerating mutations for a given interesting candidate is much cheaper
than enumerating \emph{all posible values}.
%
Hence, our tool can be used to produce much larger test cases before their size
becomes computationally too restrictive.

>>>>

There exists a vast literature on fuzzing and property-based testing.
%
To preserve space, in this section we only focus on three main topics:
%
automated random data generation using static information,
%
fuzzing using coverage information, and
%
exhaustive bounded testing tools, all of which inspired the design of \mutagen.

@@

% DRAGEN and DRAGEN2
In Haskell, DRAGEN \cite{dragen} is a meta-programming tool that synthezises
random generators from datatypes definitions, using stochastic models to predict
and optimize their distribution based on a target one set by the user.
%
DRAGEN2 \cite{dragen2} extends this idea adding support for generating richer
random data by extracting additional static information like library APIs and
function input patterns from the codebase.
%
% Bolztmann-brain
Similarly, \cite{bendkowski} developed a polinomial tuning mechanism based on
Boltzmann samplers \cite{boltzmann} that synthesizes generators with
approximate-size distributions for combinatorial structures like lists or trees.

%
% QuickFuzz
%
Exploiting some of the ideas described above, \emph{QuickFuzz} \cite{quickfuzz,
  quickfuzzjournal} is a generational grammar-based fuzzer that leverages
existing Haskell libraries describing common file formats to synthesize random
data generators that are used in tandem with off-the-shelf low-level fuzzers to
find bugs in massively used programs.

@@

We consider all these generational approaches to be orthogonal to the ideas
behind \mutagen.
%
In principle, our tool is tailored to improve the performance of poor
automatically derived generators.
%
However, more specialized generators can by used directly by \mutagen, and
combining them with well-typed mutations using a hybrid technique is an
appealing idea that we aim to address in the future.

@@

% AFL
\emph{AFL} \cite{afl} is the reference tool when it comes to coverage-guided
fuzzing.
%
% AFLfast
\emph{AFLFast} is an extension to AFL that uses Markov chain models to tune the
power scheduler towards testing low-frequency paths.
%
In \mutagen, the scheduler does not account for path frequency.
%
Instead, it favours a breadth-first traversal of the execution path space.
%
% CollAFL
Similar to our tool, \emph{CollAFL} \cite{collAFL} is a variant of AFL that uses
path- instead of edge-based coverage, which helps distinguishing executions more
precisely by reducing path collisions.

%AFLGo
When the source code history is available, \emph{AFLGo} \cite{aflgo} is a fuzzer
that can be targeted to exercise the specific parts of the the code affected by
recent commits in order to find potential new bugs.
%
Back to PBT, a related idea called \emph{targeted property-based testing} (TPBT)
is to use fitness functions to guide the testing efforts towards user defined
goals.
%
\emph{Target} \cite{targetedpbt} is a tool that implements TPBT using
optimization techniques like hill climbing and simulated annealing.

% Coverage-guided fuzzing + symbolic execution
%
Moreover, a popular technique is to combine coverage-guided fuzzing with ideas
borrowed from symbolic execution tools like \emph{KLEE} \cite{klee} to avoid
getting stuck in superficial paths \cite{driller, qsym}.
%
When using off-the-shelf symbolic executors, this technique is often limited by
the path explosion problem, altough recent tools like \emph{QSYM} \cite{qsym}
demonstrate that using tailored concolic executors can help overcoming this
limitation.

Most of these ideas can potentially be incorporated in our tool, and we keep
them as a challenge for future work.

@@

%
% Feat
%
\emph{Feat} \cite{feat} formalizes the notion of \emph{functional enumerations}.
%
For any algebraic type, it synthesizes a bijection between a finite prefix of
the natural numbers and set of increasingly growing values of the input type.
%
Later, this bijection can be traversed exhaustively or, more interestingly,
randomly accessed.
%
This allows the user to easily generate random values in a uniform basis simply
by sampling natural numbers.
%
However, values are enumerated based only on their type definition, so this
technique is not suitable for testing properties with sparse preconditions
expressed elsewhere.

% SmallCheck and LazySmallCheck
\emph{SmallCheck} \cite{smallcheck} is a Haskell tool that also follows this
idea.
%
It progressively executes the testing properties against all possible inputs
values up to a certain size bound.
%
Later, \emph{LazySmallCheck} is a variant of \emph{SmallCheck} that uses lazy
evaluation to automatically prune the search space by detecting which
subexpressions of the generated values are not evaluated by the testing
properties, and hence do not need to be populated exhaustively with concrete
values.

@@

Being exhaustive, these approaches can only be used to enumerate and test a
relatively little space of small inputs.
%
In this work we use exhaustiveness as a way to reliably enforce that all
possible mutants of an interesting seed are executed.
%
In contrast to fully-exhaustive testing tools, \mutagen relies initially on
randomness to find initial interesting mutable candidates.
%
Later, enumerating mutations for a given interesting candidate is much cheaper
than enumerating \emph{all posible values}.
%
Hence, our tool can be used to produce much larger test cases before their size
becomes computationally too restrictive.
